Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-19 paper.tex
ENH:  Finished first draft.

 \usepackage{makecell}

@@

@@

\title[Explicit B-spline regularization]{Explicit B-spline regularization in diffeomorphic image registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

As alluded to earlier, B-spline approximation can also be used
for regularizing vector fields in an analogous fashion as
Gaussian convolution.  In this vein and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to
represent the characteristic velocity fields.  However, we use the
directly manipulated free-form deformation formulation to improve the solution convergence and actual results.  This also facilitates modeling
temporal periodicity and the enforcement of stationary boundaries.

Both this latter work \citep{tustison2012a} and our earlier work
\citep{tustison2009} demonstrate that our DMFFD framework is potentially
applicable to the
entire gamut of diffeomorphic registration algorithms and provides
unique ``flavors'' of smoothing possibilities.  However, differences
between the two approaches produce characteristically different
solutions.  In addition to
smoothing kernel differences, Gaussian
convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD
approach.  Also, whereas
convolution operates entirely within the discrete space,
the fitting process constructs a continuous object prior
to any discrete sampling.  Further elaboration on these differences
are discussed in subsequent sections.

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1TimeVaryingVelocityFieldTransform.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1TimeVaryingBSplineVelocityFieldTransform.html
\end{itemize}
},
DARTEL%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianExponentialDiffeomorphicTransform.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineExponentialDiffeomorphicTransform.html
\end{itemize}
},
 and SyN%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1SyNImageRegistrationMethod.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSyNImageRegistrationMethod.html
\end{itemize}
})
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the symmetric normalization (SyN) algorithm, our focus in this work is
its B-spline equivalent which we denote as ``B-spline SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors \citep{tustison2012}).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  In this variant, the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  The algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

\begin{algorithm}
\caption{SyN algorithmic overview}
\label{alg:syn}
\begin{algorithmic}
\State $\phi_i \leftarrow \mathbf{Id}$, $\phi_i^{-1} \leftarrow \mathbf{Id}$
\Comment $i \in \{1,2\}$
\State $n \leftarrow 1$
\While {not converged}
  \State $v_1^n \leftarrow \nabla \Pi_{\sim} \left(I\circ\phi_1^{n-1},J\circ\phi_2^{n-1}\right)$
  \State $v_2^n \leftarrow \nabla \Pi_{\sim} \left(J\circ\phi_2^{n-1},I\circ\phi_1^{n-1}\right)$
  \State $v_i^n \leftarrow S_v( v_i^n )$      \Comment $S_v$ is a smoothing operation on the update transform field
  \State $\phi_i^n \leftarrow S_\phi( v_i^n \circ \phi_i^{n-1} )$      \Comment $S_\phi$ is a smoothing operation on the total transform field
  \State $\left(\phi_i^n\right)^{-1} \leftarrow Inv\left(\phi_i^n, \left(\phi_i^{n-1}\right)^{-1}\right)$
    \Comment Inverse field estimation described in \cite{avants2008}
  \State $n \leftarrow n + 1$
\EndWhile
\State \Return $\phi \leftarrow \phi_1 \circ \phi_2^{-1}$, $\phi^{-1} \leftarrow \phi_2 \circ \phi_1^{-1}$
\end{algorithmic}
\end{algorithm}

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.
Extending the work presented in \cite{tustison2009,tustison2012},
the DMFFD approach is a viable and practical alternative
for explicit regularization in diffeomorphic image registration.

As we mentioned previously, it has been shown that the preconditioned
gradient of the B-spline object is equivalent to a B-spline approximation
smoothing operation on the spatial derivative of the similarity metric
gradient \citep{tustison2009}. In the case of
$d$-dimensional image registration, the time-dependent velocity field can be
represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field and $B(\cdot)$ are the univariate B-spline
basis functions separately modulating regularity in the solution for each parametric dimension.  The preconditioned gradient analog for updating the velocity
field is as follows \citep{tustison2012}:
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi_\sim$,
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which is a slight modification of Eqn. (21) in \cite{tustison2009}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
$r$ is the spline order in all dimensions%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $c$ indexes the spatio-temporal dense metric gradient sample.
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.

@@

Although various methods exist for solving Eqns. (\ref{eq:ode}) and (\ref{eq:integral}),
we use $4^{th}$-order Runge-Kutta, i.e.
\begin{align}
  \phi_{n+1} &= \phi_{n} + \frac{1}{6}\left( k_1 + 2k_2 + 2k_3 + k_4 \right) \\
  t_{n+1} &= t_{n} + \Delta t \\
  \phi_0 &= \phi(t_0)
\end{align}
where
\begin{align}
  k_1 &= v\left( \phi_{n}, t_{n} \right)\Delta t \\
  k_2 &= v\left( \phi_{n} + \frac{k_1}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_3 &= v\left( \phi_{n} + \frac{k_2}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_4 &= v\left( \phi_{n} + k_3, t_{n} + \Delta t \right)\Delta t
\end{align}
which provides a more stable and reliable alternative than other numerical
methods \citep{press2007}.

@@

We take inspiration from the well-known
Klein comparative study in which 14 image registration
algorithms were evaluated based on performance
on publicly available labeled brain data \citep{klein2009}.
Based on the final rankings, the SyN algorithm is one
of the top-performing registration algorithms.  Its relative
performance coupled with its public availability has made
it a de facto standard for algorithmic comparison.

For our evaluation, we used the same data as that in the
Klein study.  Specifically, we used the data sets denoted as:
\begin{itemize}
  \item MGH10
  \item CUMC12
  \item IBSR18
  \item LPBA40
\end{itemize}
which are available for download from Arno Klein's website.%
\footnote{
http://mindboggle.info/papers/evaluation\_NeuroImage2009.php
}
We also use the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} which we denote as MICCAI35.  This T1-weighted
MRI data set consists of 35 subject MRI taken from the
Oasis database%
\footnote{
http://www.oasis-brains.org
}.  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allCUMC.pdf}
  \caption{}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allIBSR.pdf}
  \caption{Currently redoing on the cluster.}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allLPBA.pdf}
  \caption{}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allMGH.pdf}
  \caption{Currently redoing on the cluster.}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allMICCAI.pdf}
  \caption{}
\end{figure}

>>>>

@@

\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

@@

\title[DMFFD Diffeomorphic Image Registration]{Explicit B-spline regularization in diffeomorphic image registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

As alluded to earlier, B-spline approximation can also be used
for regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to
represent the characteristic velocity fields.  However, we use the
directly manipulated free-form deformation formulation to improve the solution convergence.  This also facilitates modeling
temporal periodicity and the enforcement of stationary boundaries.
Both this work and our earlier work
\citep{tustison2009} demonstrate that the DMFFD framework is potentially
applicable to the
entire gamut of diffeomorphic registration algorithms and provides
unique ``flavors'' of smoothing possibilities.  However, differences
between the two approaches produce characteristically different
solutions.  In addition to
smoothing kernel differences, Gaussian
convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD
approach.  Also, whereas
convolution operates entirely within the discrete space,
the fitting process constructs a continuous object prior
to any discrete sampling.  
Interestingly, a similar comparison was made with respect to 
Gaussian derivative estimation  \citep{bourma2007}.  Although 
typically estimated using  truncated, discrete Gaussian 
convolution, an alternative based on 
B-spline approximation demonstrated superior performance with
similar computational cost.  
%Further elaboration on these differences
%are discussed in subsequent sections.

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM,
DARTEL,
 and SyN)
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the symmetric normalization (SyN) algorithm, our focus in this work is
its B-spline equivalent which we denote as ``B-spline SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  In this variant, the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

\begin{algorithm}
\caption{Greedy SyN algorithm}
\label{alg:syn}
\begin{algorithmic}
\State $\phi_i \leftarrow \mathbf{Id}$, $\phi_i^{-1} \leftarrow \mathbf{Id}$
\Comment $i \in \{1,2\}$
\State $n \leftarrow 1$
\While {not converged}
  \State $v_1^n \leftarrow \nabla \Pi_{\sim} \left(I\circ\phi_1^{n-1},J\circ\phi_2^{n-1}\right)$
  \State $v_2^n \leftarrow \nabla \Pi_{\sim} \left(J\circ\phi_2^{n-1},I\circ\phi_1^{n-1}\right)$
  \State $v_i^n \leftarrow S_v( v_i^n )$      \Comment $S_v$ is a smoothing operation on the update transform field
  \State $\phi_i^n \leftarrow S_\phi( v_i^n \circ \phi_i^{n-1} )$      \Comment $S_\phi$ is a smoothing operation on the total transform field
  \State $\left(\phi_i^n\right)^{-1} \leftarrow Inv\left(\phi_i^n, \left(\phi_i^{n-1}\right)^{-1}\right)$
    \Comment Inverse field estimation described in \cite{avants2008}
  \State $n \leftarrow n + 1$
\EndWhile
\State \Return $\phi \leftarrow \phi_1 \circ \phi_2^{-1}$, $\phi^{-1} \leftarrow \phi_2 \circ \phi_1^{-1}$
\end{algorithmic}
\end{algorithm}

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  A viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient indexed by $c$.

Similarly, in the case of $d$-dimensional time-parameterized 
diffeomorphic image registration, 
the time-dependent velocity field can be represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field.  The preconditioned gradient analog of
Equation (\ref{eq:dmffd}) for updating the time-varying velocity field control point lattice
is as follows:
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi_\sim$,
\begin{align}
\label{eq:tvdmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
\footnote{
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries for all described
DMFFD-based diffeomorphic algorithms, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.
}

For regularization of constant velocity fields, e.g. SyN or DARTEL, updating the 
field is performed using Equation (\ref{eq:dmffd}).  In the case B-spline SyN, 
this applies to the smoothing operators $S_{v}$ and $S_{\phi}$ in Algorithm \ref{alg:syn}
although best performance (at least for the data described in this work) typically
employs no smoothing on the total transform field, i.e. $S_\phi$ is such that $S_\phi( \phi_i^n ) = \phi_i^n$.

@@

Three diffeomorphic algorithms described previously utilizing Gaussian convolution smoothing and their DMFFD counterparts are available through the following classes in the ITK repository: 
\begin{itemize}
  \item {\bf LDDMM:}
    \begin{itemize}
      \item {\tt itk::TimeVaryingVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::TimeVaryingVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::GaussianSmoothingOnUpdateTimeVaryingVelocityFieldTransform}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf DARTEL:}
    \begin{itemize}
       \item{\tt itk::ConstantVelocityFieldTransform}
       \item{\tt itk::ConstantVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::GaussianExponentialDiffeomorphicTransform} 
      \item {\tt itk::GaussianExponentialDiffeomorphicTransformParametersAdaptor}
      \item {\tt itk::BSplineExponentialDiffeomorphicTransform} 
      \item {\tt itk::BSplineExponentialDiffeomorphicTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf SyN:}
    \begin{itemize}
      \item {\tt itk::SyNImageRegistrationMethod} 
      \item {\tt itk::BSplineSyNImageRegistrationMethod} 
    \end{itemize}  
\end{itemize}
These and other classes (e.g. similarity metrics, optimization methods, and
utility classes) were developed as part of the ITKv4 registration
framework refactoring.  Further details can be found in the documentation
provided within the classes themselves.%
\footnote{
http://www.itk.org/Doxygen/html/classes.html
}

The class {\tt itk::BSplineScatteredDataPointSetToImageFilter} underlies all DMFFD regularization which implements the 
methods described in \cite{tustison2009}.  Although applicable to
various scenarios (e.g. curve and surface estimation), it has
been optimized for imaging applications and multi-threaded
for fast processing on multi-core machines.
Additionally, numerical integration for solving Equations (\ref{eq:ode}) and (\ref{eq:integral}) utilizes Runge-Kutta which
provides a more stable alternative than other
methods \citep{press2007}.  Implementation is provided in the class {\tt itk::TimeVaryingVelocityFieldIntegrationImageFilter}.

A complete packaging of these classes has been made
available as part of our Advanced Normalization Tools (ANTs) toolkit.%
\footnote{
http://stnava.github.io/ANTs/
}
The \verb#antsRegistration# program%
\footnote{
The long command and short command line help menus can be invoked by
the commands `{\tt antsRegistrationantsRegistration --help}' and `{\tt antsRegistration -h}', respectively.
}
takes advantage of the enhanced ITKv4
registration framework and was developed by the 
authors to provide a robust and versatile solution for a wide variety of
image registration applications. The basic conceptualization for use is that one 
can
set-up any number of registration ``stages'' with each stage being 
characterized by a specified transform.  For example, a representative 
command call is as follows:
\vspace{2mm}
\lstset{frame = htb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=bash,
        floatplacement=!h,
        caption={Representative script containing {\tt antsRegistration} and {\tt antsApplyTransforms} command calls used for evaluation},
        captionpos=b,
        label=listing:script
        }
\begin{lstlisting}
# Register the $fixed and $moving images with initial
# alignment of the centers of intensity followed by 
# the following three stages:
#   rigid -> affine -> B-spline SyN

antsRegistration --dimensionality 3 \
                 --output ${prefix} \
                 --use-histogram-matching 1 \
                 --initial-moving-transform [${fixed},${moving},1] \
                 --transform Rigid[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250 \
                 --smoothing-sigmas 2.0x1.0x0.0mm \
                 --shrink-factors 4x3x2 \
                 --transform Affine[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250 \
                 --smoothing-sigmas 2.0x1.0x0.0mm \
                 --shrink-factors 4x3x2 \
                 --transform BSplineSyN[0.1,10x10x10,0x0x0,3] \
                 --metric CC[${fixed},${moving},1,4] \ 
                 --convergence 100x70x50x20 \
                 --smoothing-sigmas 2x1.0x0.5x0.0mm \
                 --shrink-factors 6x4x2x1

# Apply the resulting transforms (generic affine + 
# B-spline SyN) to the moving labels.

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass. The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations perl resolution level) and metric (or
metrics).  Note that any pair of images can be specified per stage.

Although the
resulting transforms for each stage can be written to disk as output, the 
default output consists of a condensed set of transforms where compatible 
transforms have been composed to a single transform.  For example, in the
above command call, the initial translation, rigid, and affine transforms are
combined into a single generic affine transform file with the results of the
deformable transform consisting of discrete vector fields.  The output
transform files can then be applied using the {\tt antsApplyTransforms}
program which permits composition of any number of transform files with 
various interpolation schemes.   For both programs interpolation is never
performed more than once.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{templates.pdf}
  \caption{Canonical views for each of the five cohort-specific templates generated using
  the ANTs tools as described
  in \cite{avants2010}.   The pseudo-geodesic transform between subjects is created from 
  the composition of transforms to/from the relevant template.
  }
  \label{fig:templates}
\end{figure}

In the well-known Klein comparative study \citep{klein2009}, 14 image registration
algorithms were evaluated based on performance on publicly available labeled brain data.
%Based on the final rankings, the SyN algorithm was determined to be one
%of the top-performing deformable registration algorithms.  Its relative
%performance coupled with its public availability has made
%it a de facto standard for algorithmic comparison.
For our evaluation, we used the same data as that in the
Klein study.  Specifically, we used the data sets denoted as:
\begin{itemize}
  \item MGH10 
  \item CUMC12
  \item IBSR18
  \item LPBA40
\end{itemize}
which are available for download from Arno Klein's website.%
\footnote{
http://mindboggle.info/papers/evaluation\_NeuroImage2009.php
}
The number of subjects per cohort is provided in the denotation.
Further details of these labeled brain data (e.g. labeling protocol,
data sources) are given in \citep{klein2009}.
We also include the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} 
which we denote as MAL35.  This T1-weighted
MRI data set consists of 35 subject MRIs taken from the
Oasis database.%
\footnote{
http://www.oasis-brains.org
}  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pseudo_geodesic.pdf}
  \caption{Illustration of generating a pseudo-geodesic for any two subjects
  within the MGH10 cohort.  Once the transforms between the
  template and each subject are calculated, the mapping between any two
  subjects is found by composition of forward and inverse transforms.  For
  example, in the MGH data set, the pseudo-geodesic transform to
  map Subject g4 to Subject g7 is found by composing the forward
  transform from g4 to the template with the inverse transform from the
  template to g7 (green dashed lines).
  }
  \label{fig:pseudo_geodesic}
\end{figure}

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  This was accomplished using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{labels.pdf}
  \caption{Axial views of sample labelings for a member of each data set.  
  The second row consists of the original labelings with the third row being
  refined versions of those labelings using the MALF algorithm \cite{wang2012}.
  These refinements provide more consistency between labelings and improved
  comparative assessments between algorithms.
  }
  \label{fig:labels}
\end{figure}

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with the ANTs repository.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality bias} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{ccccc}
  {} &
  \multicolumn{2}{c}{\bf Original labeling} & 
  \multicolumn{2}{c}{\bf MALF labeling} \\
  \rotatebox{90}{\qquad\,\, CUMC12} &
  \includegraphics[height=38mm]{versusCUMC.png} &
  \includegraphics[height=38mm]{synComparisonCUMC.pdf} &
  \includegraphics[height=38mm]{versusCUMCMALF.png} &
  \includegraphics[height=38mm]{synComparisonCUMCMALF.pdf} \\
  \rotatebox{90}{\qquad\quad IBSR18} &
  \includegraphics[height=38mm]{versusIBSR.png} &
  \includegraphics[height=38mm]{synComparisonIBSR.pdf} &
  \includegraphics[height=38mm]{versusIBSRMALF.png} &
  \includegraphics[height=38mm]{synComparisonIBSRMALF.pdf} \\
  \rotatebox{90}{\qquad\quad LPBA40} &
  \includegraphics[height=38mm]{versusLPBA.png} &
  \includegraphics[height=38mm]{synComparisonLPBA.pdf} &
  \includegraphics[height=38mm]{versusLPBAMALF.png} &
  \includegraphics[height=38mm]{synComparisonLPBAMALF.pdf} \\
  \rotatebox{90}{\qquad\quad MGH10} &
  \includegraphics[height=38mm]{versusMGH.png} &
  \includegraphics[height=38mm]{synComparisonMGH.pdf} &
  \includegraphics[height=38mm]{versusMGHMALF.png} &
  \includegraphics[height=38mm]{synComparisonMGHMALF.pdf} \\
  \rotatebox{90}{\qquad\quad MAL35} &
  \includegraphics[height=38mm]{versusMAL.png} &
  \includegraphics[height=38mm]{synComparisonMAL.pdf} &
  \includegraphics[height=38mm]{versusMALMALF.png} &
  \includegraphics[height=38mm]{synComparisonMALMALF.pdf} \\
  \end{tabular}
  \caption{Dice results for both algorithms for 
  each subject warped to every other subject using the pseudo-
  geodesic transform.  Each 
  row corresponds to one of the five data sets used for evaluation.
  For each data set we include a plotting of all individual label 
  Dice results by volume and a combined label box plotting.  The
  left and right halves show the results for MALF-derived and 
  original labelings, respectively.
  }
  \label{fig:dice}
\end{figure}

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  Note that parameter
exploration is not a part of this work.  Rather, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 during each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ with Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
All perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
It should be noted that for the
data sets used in this study, times for B-spline SyN are approximately 
15-40\% greater than Gaussian-based SyN (single-threaded).
